Mixing Time Estimation in Reversible Markov Chains from a Single Sample Path

### 2015-06-09 nips2015/notation.tex
reorganize files

%!TEX root =  matrix-est.tex

\subsection{Notations}
\label{sec:notation}

We denote the set of positive integers by $\bbN$, and
the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$,
$\log$ denotes the natural logarithm.
Boldface symbols are used for vectors and matrices (e.g., $\vv$,
$\vM$), and their entries are referenced by subindexing (e.g., $v_i$,
$M_{i,j}$).
%The $i$-th coordinate basis vector is denoted by $\ve_i$.
For a vector $\vv$, $\norm{\vv}$ denotes its Euclidean norm; for a
matrix $\vM$, $\norm{\vM}$ denotes its spectral norm.
We use $\Diag(\vv)$ to denote the diagonal matrix whose $(i,i)$-th
entry is $v_i$.
The probability simplex is denoted by $\Delta^{d-1} = \{ \vp
\in [0,1]^d : \sum_{i=1}^d p_i = 1 \}$, and we regard vectors in
$\Delta^{d-1}$ as row vectors.
%Finally, the total variation between two probability distributions
%$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.

\subsection{Setting}
\label{sec:setting}

Let $\vP \in (\Delta^{d-1})^d \subset [0,1]^{d \times d}$ be a $d
\times d$ row-stochastic matrix for an ergodic (i.e., irreducible and
aperiodic) Markov chain.
This implies there is a unique stationary distribution $\vpi \in
\Delta^{d-1}$ with $\pi_i > 0$ for all $i \in [d]$~\citep[Corollary
1.17]{LePeWi08}.
We also assume that $\vP$ is \emph{reversible} (with respect to
$\vpi$):
\begin{align}
\label{eq:reversibility}
  \pi_i P_{i,j} = \pi_j P_{j,i} ,
  \quad i,j \in [d] .
\end{align}
The minimum stationary probability is denoted by $\pimin := \min_{i
\in [d]} \pi_i$.

Define the matrices
\begin{align*}
\vM := \Diag(\vpi) \vP \quad \text{and} \quad
\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}\,.
\end{align*}
The $(i,j)$th entry of the matrix $M_{i,j}$ contains the \emph{doublet
probabilities} associated with $\vP$: $M_{i,j} = \pi_i P_{i,j}$ is the
probability of seeing state $i$ followed by state $j$ when the chain
is started from its stationary distribution.
The matrix $\vM$ is symmetric on account of the reversibility of
$\vP$, and hence it follows that $\vL$ is also symmetric.
(We will strongly exploit the symmetry in our results.)
%will be advantageous due to the existence of stronger perturbation results
%for symmetric matrices.
Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
and $\vP$ are similar and thus their eigenvalue systems are identical.
Ergodicity and reversibility imply that the eigenvalues of $\vL$ are
contained in the interval $\opencloseint{-1,1}$, and that $1$ is an
eigenvalue of $\vL$ with multiplicity $1$~\citep[Lemmas 12.1 and
12.2]{LePeWi08}.
Denote and order the eigenvalues of $\vL$ as
\[
  1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1 .
\]
Let $\slem := \max\{ \lambda_2,\, |\lambda_d| \}$, and define the
(absolute) spectral gap to be $\gap := 1-\slem$, which is strictly
positive on account of ergodicity.
\todoc{Do our results generalize to the spectral gap? No.
Where do we need that we talk about the absolute spectral gap?
What is the difference between the spectral and the absolute spectral gap? Say, in their relationship to $\tmix$?}
%Throughout the paper, both $\pimin$ and $\gap$ are assumed
%to be positive.

Let $(X_t)_{t\in\bbN}$ be a Markov chain whose transition
probabilities are governed by $\vP$.
For each $t \in \bbN$, let $\vpi^{(t)} \in \Delta^{d-1}$ denote the
marginal distribution of $X_t$, so
\[
  \vpi^{(t+1)} = \vpi^{(t)} \vP ,
  \quad t \in \bbN .
\]
Note that the initial distribution $\vpi^{(1)}$ is arbitrary,
and need not be the stationary distribution $\vpi$.

The goal is to estimate $\pimin$ and $\gap$ from the length $n$ sample
path $(X_t)_{t \in [n]}$, and also to construct fully empirical
confidence intervals that $\pimin$ and $\gap$ with high probability;
in particular, the construction of the intervals should not depend on
any unobservable quantities, including $\pimin$ and $\gap$ themselves.
As mentioned in the introduction,
it is well-known that the \emph{mixing time} of the Markov chain
$\tmix$ (defined in Eq.~\ref{eq:mixtimedef})
is bounded in terms of $\pimin$ and $\gap$, as shown in
\cref{eq:mixing-time-bound}.
Moreover, convergence rates for empirical processes on Markov chain
sequences are also often given in terms of mixing coefficients that
can ultimately be bounded in terms of $\pimin$ and
$\gap$ (as we will show in the proof of our first result).
%~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 
Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
used to make these rates fully observable.
All probabilities in this paper are over the space of state sequences with the given Markov measure.
\todoa{Added this last sentence}

%Therefore, confidence intervals for $\pimin$ and $\gap$ can be used to
%make these rates data-dependent and hence usable in procedures such
%as Structural Risk Minimization.
%\todoc{citation for SRM, and other applications?}
%
%Additionally, we aim to provide
%asymptotic convergence rates
%as well as data-dependent confidence intervals;
%the latter should not depend on any unknown quantities
%(such as $\pimin$ and $\gap$).
%
%\todoc[inline]{What is the data? What is the problem?}

>>>>

%!TEX root =  matrix-est.tex
We denote the set of positive integers by $\bbN$, and
the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$.
We will use $\ln(x)$, to denote the natural logarithm of $x$
in expressions with exact numerical constants,
while we use $\log(x)$ elsewhere.
Boldface symbols are used for vectors and matrices (e.g., $\vv$,
$\vM$), and their entries are referenced by subindexing (e.g., $v_i$,
$M_{i,j}$).
%The $i$-th coordinate basis vector is denoted by $\ve_i$.
For a vector $\vv$, $\norm{\vv}$ denotes its Euclidean norm; for a
matrix $\vM$, $\norm{\vM}$ denotes its spectral norm.
We use $\Diag(\vv)$ to denote the diagonal matrix whose $(i,i)$-th
entry is $v_i$.
The probability simplex is denoted by $\Delta^{d-1} = \{ \vp
\in [0,1]^d : \sum_{i=1}^d p_i = 1 \}$, and we regard vectors in
$\Delta^{d-1}$ as row vectors.
%Finally, the total variation between two probability distributions
%$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.

### 2015-06-09 nips2015/setting.tex
reorganize files

The goal is to estimate $\pimin$ and $\gap$ from the length $n$ sample
path $(X_t)_{t \in [n]}$, and also to construct fully empirical
confidence intervals that $\pimin$ and $\gap$ with high probability;
in particular, the construction of the intervals should not depend on
any unobservable quantities, including $\pimin$ and $\gap$ themselves.
As mentioned in the introduction,
it is well-known that the \emph{mixing time} of the Markov chain
$\tmix$ (defined in Eq.~\ref{eq:mixtimedef})
is bounded in terms of $\pimin$ and $\gap$, as shown in
\cref{eq:mixing-time-bound}.
Moreover, convergence rates for empirical processes on Markov chain
sequences are also often given in terms of mixing coefficients that
can ultimately be bounded in terms of $\pimin$ and
$\gap$ (as we will show in the proof of our first result).
%~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 
Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
used to make these rates fully observable.
All probabilities in this paper are over the space of state sequences with the given Markov measure.
\todoa{Added this last sentence}

>>>>

The goal is to estimate $\pimin$ and $\gap$ from the length $n$ sample
path $(X_t)_{t \in [n]}$, and also to construct fully empirical
confidence intervals that $\pimin$ and $\gap$ with high probability;
in particular, the construction of the intervals should not depend on
any unobservable quantities, including $\pimin$ and $\gap$ themselves.
As mentioned in the introduction,
it is well-known that the \emph{mixing time} of the Markov chain
$\tmix$ (defined in Eq.~\ref{eq:mixtimedef})
is bounded in terms of $\pimin$ and $\gap$, as shown in
\cref{eq:mixing-time-bound}.
Moreover, convergence rates for empirical processes on Markov chain
sequences are also often given in terms of mixing coefficients that
can ultimately be bounded in terms of $\pimin$ and
$\gap$ (as we will show in the proof of our first result).
%~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 
Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
used to make these rates fully observable.
%All probabilities in this paper are over the space of state sequences with the given Markov measure.
\todoa{Add: ``All probabilities in this paper are over the space of state sequences with the given Markov measure.'' 
Cs: I think the probability space could be arbitrary
and the probabilities are over this (implicit) probability space as usual in probability papers.
I would suggest to remove this sentence. Also, we are short of space.}
